Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-14 response-to-reviewers.tex

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
stuff
\end{point}

\begin{reply}
responds
\end{reply}

\reviewersection

\begin{point}
stuff
\end{point}

\begin{reply}
responds
\end{reply}

\end{document}

### 2021-11-14 reviewed-paper.tex
Infrastructure for latexdiff

% Middle Authors
\author[8]{Bjarki Eldon}
\author[9]{E. Castedo Ellerman}
\author[10,11]{Jared G. Galloway}
\author[12,13]{Ariella L. Gladstein}
\author[14]{Gregor Gorjanc}
\author[15]{Bing Guo}
\author[7]{Ben Jeffery}
\author[16]{Warren W. Kretzschmar}
% Haven't been able to contact Ivan, so forced to drop for preprint
% \author[17]{Ivan Krukov}
\author[2]{Konrad Lohse}
\author[17]{Michael Matschiner}
\author[18]{Dominic Nelson}
\author[19]{Nathaniel S. Pope}
\author[20]{Consuelo D. Quinto-Cort\'es}
\author[10]{Murillo F. Rodrigues}
\author[21]{Kumar Saunack}
\author[22]{Thibaut Sellinger}
\author[23]{Kevin Thornton}
\author[9]{Hugo van Kemenade}
\author[7,24]{Anthony W. Wohns}
\author[7]{H. Yan Wong}

@@

Recombination events have a dominant effect on the execution time of
simulations. For details on the time complexity of the simulation algorithm,
see the appendix: \nameref{app-time-complexity}.

@@

We have added two multiple merger coalescent models in \msprime\ 1.0, the
Beta-coalescent~\citep{schweinsberg03} and
``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
such models with recombination for the first time.
These simulation models have been extensively validated against
analytical results from the site frequency
spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
 as well as more general properties of coalescent processes.
See the \nameref{app-multiple-mergers} appendix for more details and model derivations.

@@

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the chromosome,
and the ancestral genomes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral genomes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and chromosomal location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which it is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence
(Fig.~\ref{fig-arg}).
One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
and mutational patterns under
the standard coalescent and infinite sites mutation model.
For details, see the appendix: \nameref{app-likelihoods}.

@@

\subsection*{Time complexity of the algorithm}
\label{app-time-complexity}

As discussed in the \nameref{sec-recombination} section, the time complexity of
Hudson's algorithm is quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of
recombination distance. For large $\rho$, recombination events in trapped
ancestral material will dominate, and so we can use this as a proxy for the
overall number of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.

\subsubsection*{Practical examples}

\subsubsection*{Changing population size}

@@

\subsection*{Multiple merger coalescent model}
\label{app-multiple-mergers}

>>>>

% Middle Authors
\author[8]{Bjarki Eldon}
\author[9]{Castedo E. Ellerman}
\author[10,11]{Jared G. Galloway}
\author[12,13]{Ariella L. Gladstein}
\author[14]{Gregor Gorjanc}
\author[15]{Bing Guo}
\author[7]{Ben Jeffery}
\author[16]{Warren W. Kretzschmar}
% Haven't been able to contact Ivan, so forced to drop for preprint
% \author[17]{Ivan Krukov}
\author[2]{Konrad Lohse}
\author[17]{Michael Matschiner}
\author[18]{Dominic Nelson}
\author[19]{Nathaniel S. Pope}
\author[20]{Consuelo D. Quinto-Cort\'es}
\author[10]{Murillo F. Rodrigues}
\author[21]{Kumar Saunack}
\author[22]{Thibaut Sellinger}
\author[23]{Kevin Thornton}
\author[9]{Hugo van Kemenade}
\author[7,24]{Anthony W. Wohns}
\author[7]{H. Yan Wong}

@@

For large $\rho$, recombination events in trapped ancestral material will
dominate, and so we can use this as a proxy for the overall number
of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.

@@

We have added two multiple merger coalescent models in \msprime\ 1.0, the
Beta-coalescent~\citep{schweinsberg03} and
``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
such models with recombination for the first time.
These simulation models have been extensively validated against
analytical results from the site frequency
spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
 as well as more general properties of coalescent processes.
See the Appendix for more details and model derivations.

@@

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the chromosome,
and the ancestral genomes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral genomes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and chromosomal location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which it is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence
(Fig.~\ref{fig-arg}).
One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details on these likelihood calculations.

@@

@@

\label{app-multiple-mergers}
\subsection*{Multiple merger coalescent model}
